---
title: "Programming Basics Exercises (Lecture 6)"
author: | 
  | W. Evan Johnson, Ph.D.
  | Professor, Division of Infectious Disease
  | Director, Center for Data Science
  | Rutgers University -- New Jersey Medical School
date: "4/20/2024"
output: pdf_document

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These exercises will give you some introductory experience with programming basics. Please complete the following: 

1\. What will this conditional expression return?

```{r, eval=FALSE}
x <- c(1,2,-3,4)

if(all(x>0)){
  print("All Postives")
} else{
  print("Not all positives")
}
```


2\. Which of the following expressions is always `FALSE` when at least one entry of a logical vector `x` is TRUE?

a. `all(x)`
b. `any(x)`
c. `any(!x)`
d. `all(!x)`

3\. The function `nchar` tells you how many characters long a character vector is. Write a line of code that assigns to the object `new_names` the state abbreviation when the state name is longer than 8 characters.
    
  
4\. Create a function `sum_n` that for any given value, say $n$, computes the sum of the integers from 1 to n (inclusive). Use the function to determine the sum of integers from 1 to 5,000.

5\. Create a function `altman_plot` that takes two arguments, `x` and `y`, and plots the difference against the sum.
  
6\. After running the code below, what is the value of `x`?

```{r, eval=FALSE}
x <- 3
my_func <- function(y){
  x <- 5
  y+5
}
```

7\. Write a function `compute_s_n` that for any given $n$ computes the sum $S_n = 1^2 + 2^2 + 3^2 + \dots n^2$. Report the value of the sum when $n=10$.
 
8\. Define an empty numerical vector `s_n` of size 25 using `s_n <- vector("numeric", 25)` and store in the results of $S_1, S_2, \dots S_{25}$ using a for-loop.

9\. Repeat exercise 8, but this time use `sapply`.

10\. Repeat exercise 8, but this time use `map_dbl`.

11\. Plot $S_n$ versus $n$. Use points defined by $n=1,\dots,25$.

12\. Confirm that the formula for this sum is $S_n= n(n+1)(2n+1)/6$.

