---
title: "Lecture 4: R Basics, Part 2"
author: | 
  | W. Evan Johnson, Ph.D.
  | Professor, Division of Infectious Disease
  | Director, Center for Data Science
  | Rutgers University -- New Jersey Medical School
date: "4/20/2024"
header-includes:
   - \usepackage{amsmath}
output: 
  beamer_presentation:
    theme: "Warsaw"
editor_options: 
  chunk_output_type: console
tables: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
img_path <- "img/"
library(tidyverse)
library(ggplot2)
library(ggflags)
library(countrycode)
```

## Data Types

Variables in R can be of different types. For example: numbers, character strings, tables simple lists. The function `class` helps us determine what type of object we have:

```{r}
a <- 2
class(a)
```

To work efficiently in R, it is important to learn the different types of variables and what we can do with these.

## Data Frames 

Up to now, the variables we have defined are just one number, which is not very useful for storing data. The most common way of storing a dataset in R is in a __data frame__. 

Conceptually, we can think of a data frame as a table with rows representing observations and the different variables reported for each observation defining the columns. 

Data frames are particularly useful for datasets because we can combine different data types into one object. 

## Data Frames 
A large proportion of data analysis challenges start with data stored in a data frame. For example, we stored the data for our motivating example in a data frame. You can access this dataset by loading the __dslabs__ library and loading the `murders` dataset using the `data` function:

```{r}
library(dslabs)
data(murders)
```

## Data Frames 
To see that this is in fact a data frame, we type:

```{r}
class(murders)
```

## Examining an Object

The function `str` is useful for finding out more about the structure of an object:

```{r, echo=FALSE}
# change str to make sure it stays within margins
str <- function(x) utils::str(x, strict.width = 'wrap')
```

```{r}
str(murders)
```

```{r, echo=FALSE}
# remove str we defiend
rm(str)
```

This tells us much more about the object. We see that the table has 51 rows (50 states plus DC) and five variables. 

## Data Frames 
We can show the first six lines using the function `head`:

```{r}
head(murders)
```

## Data Frames 
In this dataset, each state is considered an observation and five variables are reported for each state.

Before we go any further in answering our original question about different states, let's learn more about the components of this object.

## The Accessor: `$`

For our analysis, we will need to access the different variables represented by columns included in this data frame. To do this, we use the accessor operator `$` in the following way:
\footnotesize
```{r}
murders$population
```

## The Accessor: `$`
But how did we know to use `population`? Previously, by applying the function `str` to the object `murders`, we revealed the names for each of the five variables stored in this table. We can quickly access the variable names using:

\footnotesize
```{r}
names(murders)
```

## The Accessor: `$`
**Important:** Note the order of the entries in `murders$population` preserves the order of the rows in our data table. This will later permit us to manipulate one variable based on the results of another. For example, we will be able to order the state names by the number of murders.

**Pro Tip**: R comes with a very nice auto-complete functionality that saves us the trouble of typing out all the names. Try typing `murders$p` then hitting the __tab__ key on your keyboard. This functionality and many other useful auto-complete features are available when working in RStudio.

## Vectors: Numerics, Characters, and Logical

The object `murders$population` is not one number but several. We call these types of objects __vectors__. A single number is technically a vector of length 1, but in general we use the term vectors to refer to objects with several entries. The function `length` tells you how many entries are in the vector:

```{r}
pop <- murders$population
length(pop)
```

## Vectors: Numerics, Characters, and Logical
This particular vector is __numeric__ since population sizes are numbers:

```{r}
class(pop)
```
In a numeric vector, every entry must be a number. 

## Vectors: Numerics, Characters, and Logical
To store character strings, vectors can also be of class __character__. For example, the state names are characters:

```{r}
class(murders$state)
```

As with numeric vectors, all entries in a character vector need to be a character.

## Vectors: Numerics, Characters, and Logical
Another important type of vectors are __logical vectors__. These must be either `TRUE` or `FALSE`. 

```{r}
z <- 3 == 2
z
class(z)
```

Here the `==` is a relational operator asking if 3 is equal to 2. In R, if you just use one `=`, you actually assign a variable, but if you use two `==` you test for equality. 

## Vectors: Numerics, Characters, and Logical
You can see the other __relational operators__ by typing:

```{r, eval=FALSE}
?Comparison
```

In the future, you will see how useful relational operators can be. We discuss more important features of vectors later.


## Vectors: Numerics, Characters, and Logical
**Advanced**: Mathematically, the values in `pop` are integers and there is an integer class in R. However, by default, numbers are assigned class numeric even when they are round integers. For example, `class(1)` returns numeric. You can turn them into class integer with the `as.integer()` function or by adding an `L` like this: `1L`. Note the class by typing: `class(1L)`


## Factors 

In the `murders` dataset, we might expect the region to also be a character vector. However, it is not:

```{r}
class(murders$region)
```

It is a __factor__. Factors are useful for storing categorical data. 


## Factors 
We can see that there 4 regions by using the `levels` function:

\footnotesize
```{r}
levels(murders$region)
```

## Factors 
In the background, R stores these __levels__ as integers and keeps a map to keep track of the labels. This is more memory efficient than storing all the characters. 

Note that the levels have an order that is different from the order of appearance in the factor object. The default in R is for the levels to follow alphabetical order. However, often we want the levels to follow a different order. 

You can specify an order through the `levels` argument when creating the factor with the `factor` function. For example, in the murders dataset regions are ordered from east to west.  The function `reorder` lets us change the order of the levels of a factor variable based on a summary computed on a numeric vector. 

## Factors 

Suppose we want the levels of the region by the total number of murders rather than alphabetical order. If there are values associated with each level, we can use the `reorder` and specify a data summary to determine the order. The following code takes the sum of the total murders in each region, and reorders the factor following these sums.

\footnotesize

```{r}
region <- murders$region
value <- murders$total
region <- reorder(region, value, FUN = sum)
levels(region)
```

## Factors 
The new order is in agreement with the fact that the Northeast has the least murders and the South has the most.

**Warning**: Factors can be a source of confusion since sometimes they behave like characters and sometimes they do not. As a result, confusing factors and characters are a common source of bugs.  

## Lists

Data frames are a special case of __lists__. Lists are useful because you can store any combination of different types. You can create a list using the `list` function like this:

```{r}
record <- list(name = "John Doe",
             student_id = 1234,
             grades = c(95, 82, 91, 97, 93),
             final_grade = "A")
```

The function `c` (for concatenate) is described later.

## Lists
The list on the prior slide includes a character, a number, a vector with five numbers, and another character.
\footnotesize
```{r}
class(record)
record
```

## Lists

As with data frames, you can extract the components of a list with the accessor `$`. 

```{r}
record$student_id
```

## Lists
We can also use double square brackets (`[[`) like this:

```{r}
record[["student_id"]]
```

You should get used to the fact that in R, there are often several ways to do the same thing, such as accessing entries.

## Lists
You might also encounter lists without variable names.

```{r,}
record2 <- list("John Doe", 1234)
record2
```

## Lists
If a list does not have names, you cannot extract the elements with `$`, but you can still use the brackets method and instead of providing the variable name, you provide the list index, like this:

```{r}
record2[[1]]
```

We won't be using lists until later, but you might encounter one in your own exploration of R. For this reason, we showed you some basics here. 


## Matrices 

__Matrices__ are another type of object that are common in R. Matrices are similar to data frames in that they are two-dimensional: they have rows and columns. However, like numeric, character and logical vectors, entries in matrices have to be all the same type. 
For this reason data frames are much more useful for storing data, since we can have characters, factors, and numbers in them. 

## Matrices 
Yet matrices have a major advantage over data frames: we can perform matrix algebra operations, a powerful type of mathematical technique. We will cover matematical computions on matrices in more detail later!

## Matrices 
We can define a matrix using the `matrix` function. We need to specify the number of rows and columns.

```{r}
mat <- matrix(1:12, 4, 3)
mat
```

## Matrices 
You can access specific entries in a matrix using square brackets (`[`). If you want the second row, third column, you use:

```{r}
mat[2, 3]
```

## Matrices 
If you want the entire second row, you leave the column spot empty:

```{r}
mat[2, ]
```

Notice that this returns a vector, not a matrix.

## Matrices 
Similarly, if you want the entire third column, you leave the row spot empty:

```{r}
mat[, 3]
```

This is also a vector, not a matrix.

## Matrices 
You can access more than one column or more than one row if you like. This will give you a new matrix.

```{r}
mat[, 2:3]
```

## Matrices 
You can subset both rows and columns:

```{r}
mat[1:2, 2:3]
```

## Matrices 
We can convert matrices into data frames using the function `as.data.frame`:

```{r}
as.data.frame(mat)
```

## Matrices 
You can also use single square brackets (`[`) to access rows and columns of a data frame:

```{r}
data("murders")
murders[25, 1]
murders[2:3, ]
```


## Exercises 
Now open the **R Basics Exercises** file and complete Exercises 6-11.  
<!--
1\. Load the US murders dataset.

```{r}
library(dslabs)
data(murders)
```

Use the function `str` to examine the structure of the `murders` object. Which of the following best describes the variables represented in this data frame?
    
a. The 51 states.
b. The murder rates for all 50 states and DC.
c. The state name, the abbreviation of the state name, the state's region, and the state's population and total number of murders for 2010. 
d. `str` shows no relevant information.
    
2\. What are the column names used by the data frame for these five variables?
    
3\. Use the accessor `$` to extract the state abbreviations and assign them to the object `a`. What is the class of this object?

4\. Now use the square brackets to extract the state abbreviations and assign them to the object `b`. Use the `identical` function to determine if `a` and `b` are the same.

5\. We saw that the `region` column stores a factor. You can corroborate this by typing: 

```{r, eval=FALSE}
class(murders$region)
```

With one line of code, use the function `levels` and `length` to determine the number of regions defined by this dataset.

6\. The function `table` takes a vector and returns the frequency of each element. You can quickly see how many states are in each region by applying this function. Use this function in one line of code to create a table of states per region.
--> 

##  Vectors 

In R, the most basic objects available to store data are __vectors__. As we have seen, complex datasets can usually be broken down into components that are vectors. For example, in a data frame, each column is a vector. Here we learn more about this important class.

## Creating Vectors

We can create vectors using the function `c`, which stands for __concatenate__. We use `c` to concatenate entries in the following way:

```{r}
codes <- c(380, 124, 818)
codes
```

## Creating Vectors
We can also create character vectors. We use the quotes to denote that the entries are characters rather than variable names.

```{r}
country <- c("italy", "canada", "egypt")
```

## Creating Vectors
In R you can also use single quotes:

```{r}
country <- c('italy', 'canada', 'egypt')
```

But be careful not to confuse the single quote ' with the __back quote__ `. 

## Creating Vectors
By now you should know that if you type:

```{r, eval=FALSE}
country <- c(italy, canada, egypt)
```
you receive an error because the variables `'italy'`, `'canada'`, and `'egypt'` are not defined. If we do not use the quotes, R looks for variables with those names and returns an error.

## Names

Sometimes it is useful to name the entries of a vector. For example, when defining a vector of country codes, we can use the names to connect the two:

```{r}
codes <- c(italy = 380, canada = 124, egypt = 818)
codes
```

## Names
The object `codes` continues to be a numeric vector:
```{r}
class(codes)
```
but with names:
```{r}
names(codes)
```

## Names
If the use of strings without quotes looks confusing, know that you can use the quotes as well:

```{r}
codes <- c("italy" = 380, "canada" = 124, "egypt" = 818)
codes
```

There is no difference between this function call and the previous one. This is one of the many ways in which R is quirky compared to other languages.

## Names
We can also assign names using the `names` functions:

```{r}
codes <- c(380, 124, 818)
country <- c("italy","canada","egypt")
names(codes) <- country
codes
```

## Sequences

Another useful function for creating vectors generates sequences:

```{r}
seq(1, 10)
```

## Sequences
The first argument defines the start, and the second defines the end which is included. The default is to go up in increments of 1, but a third argument lets us tell it how much to jump by:

```{r}
seq(1, 10, 2)
```

## Sequences
If we want consecutive integers, we can use the following shorthand:

```{r}
1:10
```

## Sequences
When we use these functions, R produces integers, not numerics, because they are typically used to index something:

```{r}
class(1:10)
```

However, if we create a sequence including non-integers, the class changes:

```{r}
class(seq(1, 10, 0.5))
```

## Subsetting

We use square brackets to access specific elements of a vector. For the vector `codes` we defined above, we can access the second element using:
```{r}
codes[2]
```

You can get more than one entry by using a multi-entry vector as an index:
```{r}
codes[c(1,3)]
```

## Subsetting
The sequences defined above are particularly useful if we want to access, say, the first two elements:

```{r}
codes[1:2]
```

## Subsetting
If the elements have names, we can also access the entries using these names. Below are two examples.

```{r}
codes["canada"]
codes[c("egypt","italy")]
```

## Coercion

In general, __coercion__ is an attempt by R to be flexible with data types. When an entry does not match the expected, some of the prebuilt R functions try to guess what was meant before throwing an error. This can also lead to confusion. 

Failing to understand __coercion__ can drive programmers crazy in R since it behaves quite differently from most other languages in this regard. Let's learn about it with some examples.

## Coercion
We said that vectors must be all of the same type. So if we try to combine, say, numbers and characters, you might expect an error:

```{r}
x <- c(1, "canada", 3)
```

But we don't get one, not even a warning! What happened? Look at `x` and its class:

```{r}
x
class(x)
```

## Coercion
R __coerced__ the data into characters. It guessed that because you put a character string in the vector, you meant the 1 and 3 to actually be character strings `"1"` and "`3`". The fact that not even a warning is issued is an example of how coercion can cause many unnoticed errors in R. 

R also offers functions to change from one type to another. For example, you can turn numbers into characters with:

```{r}
x <- 1:5
y <- as.character(x)
y
```

## Coercion
You can turn it back with `as.numeric`:

```{r}
as.numeric(y)
```
 
This function is actually quite useful since datasets that include numbers as character strings are common. 

## Not Availables (NA)

When a function tries to coerce one type to another and encounters an impossible case, it usually gives us a warning and turns the entry into a special value called an `NA` for "not available".  For example:

```{r}
x <- c("1", "b", "3")
as.numeric(x)
```

R does not have any guesses for what number you want when you type `b`, so it does not try. As a data scientist you will encounter the `NA`s often as they are generally used for missing data, a common problem in real-world datasets.

## Exercises
Now open the **R Basics Exercises** file and complete Exercises 12-23.  

<!--
1\. Use the function `c` to create a vector with the average high temperatures in January for Beijing, Lagos, Paris, Rio de Janeiro, San Juan, and Toronto, which are 35, 88, 42, 84, 81, and 30 degrees Fahrenheit. Call the object `temp`.

2\. Now create a vector with the city names and call the object `city`.


3\. Use the `names` function and the objects defined in the previous exercises to associate the temperature data with its corresponding city.

 
4\. Use the `[` and `:` operators to access the temperature of the first three cities on the list.
    
 
5\. Use the `[` operator to access the temperature of Paris and San Juan. 
   
6\. Use the `:` operator to create a sequence of numbers $12,13,14,\dots,73$. 
   
7\. Create a vector containing all the positive odd numbers smaller than 100.

8\. Create a vector of numbers that starts at 6, does not pass 55, and adds numbers in increments of 4/7: 6, 6 + 4/7, 6 + 8/7, and so on. How many numbers does the list have? Hint: use `seq` and `length`.
 
9\. What is the class of the following object `a <- seq(1, 10, 0.5)`?



10\. What is the class of the following object `a <- seq(1, 10)`?

   
11\. The class of `class(a<-1)` is numeric, not integer. R defaults to numeric and to force an integer, you need to add the letter `L`. Confirm that the class of `1L` is integer.

  
12\. Define the following vector:

```{r}
x <- c("1", "3", "5")
```

and coerce it to get integers.
-->


## Session Info
\tiny
```{r session}
sessionInfo()
```


